Community-Conditioned Language Models

### 2021-06-10 paper/paper.tex
tweaks

We also consider the average information gain per token of the CCLM over its baseline
un-conditioned counterpart, with the same sequence encoder architecture.
For a given message, information gain is defined as the difference
between the cross-entropy of the unconditioned model and the conditioned model:
\[H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m)\]
For a set of messages, $M$, we consider the average information gain
in exponential space (as a ratio of perplexities):
\[\IG_M = \frac{e^{\avg_M(H_{\mathrm{LM}}(m))}}{e^{\avg_M(H_{\mathrm{CCLM}}(m))}}\]
\[\IG_M = {e^{\avg_{m∈M}(H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m))}}\]
%
Unsurprisingly, the conditioned models mostly have lower perplexity 
than their respective unconditioned baseline models, 
meaning information gain above one (\cref{tab:model-results}).
Overall, the LSTM models perform better than the transformers, 
although the best transformer has higher information gain than the best LSTM.

@@

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-info-gain}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are \subreddit{counting}, \subreddit{friendsafari}, and \subreddit{Fireteams}, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly structured and regular formats, which 
  are presumably conventional to these communities.}

The variation in information gain across communities is not 
related to perplexity, however (\cref{fig:ppl-info-gain}).

@@

As an informal observation, we note that across all the models we
tested, communities where conditioning has the least effect tend to be
more organized around more general interest topics, such as 
\subreddit{relationships} and \subreddit{advice}, where the subject
matter is relevant to a broad range of people.  Conditioning the model
on community has the most benefit for narrower special-interest
subreddits, such as those organized around a certain videogame, sports
team, or subculture.  This makes sense intuitively, since communities
with more niche topics would tend to have more specialized vocabulary
and other linguistic patterns.

@@

We can characterize the performance of the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test set) message $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal is clearly above the rest: on average the LMCC generally
picks assigns the highest probability to the community where a message originates (\cref{fig:confusion}).

@@

But we want a more quantitative measure, corresponding to the ability of the LMCC to correctly
recognize messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use a perplexity-like metric,
which is has stronger information-theoretic grounds.
To this end, we define the \emph{linguistic indiscernibility} of community $c_j$, 
noted $\Ind_j$. 
First, note that each row of $C$ is a probability distribution,
since it is computed as the average of $P(c=c_i \mid m)$ for $m\in M_j$.
The exponential of the entropy of the $j$th row of \(C\) is therefore a perplexity score,
whose maximum is the total number of communities (46 for us).
Thus we define $\Ind_j$ by scaling this perplexity by the number of communities:
\[\Ind_j = \frac{e^{-\sum_i C_{ij} log(C_{ij})}}{|C_j|}\]
This way $Ind_j=0$ means that the LMCC is completely certain of its
choice, while $Ind_j=1$ means that equal probability is given to all
predictions.  In general $\Ind_j$ is a score of uncertainty of the
LMCC, given messages from a community $c_j$.

@@

Finally, we note that, even though community conditioning affects
perplexity very little overall, it provides large benefits in term of
the underlying Bayesian classifier. Indiscernibility is on average
\num{0.57} with a standard deviation of \num{0.22}. Roughly speaking, this means
that, while the linguistic variations does not allow our model to
pinpoint the originating community of a message, it can often
eliminate TODO (a percentatge?) of communities as possible locations \bn{43\%?}
for it. Further more, the information gain provided by
community-conditioning is a good predictor of how specific the
community is.

@@

The problem with this approach is that the dimensions of \(L\) and \(S\) may
not coincide. Applying a simple permutation of dimensions may widely
change the correlation metric. To make the metric resilient to such
effects, we can compute the minimum distance between \(L_i\) and \(S_i\), for
any orthogonal matrix \(\Omega\) applied to \(L\):

@@

Computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}. The
solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). That is, each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

Further, we can apply a permutation to \(U\), \(V\) and \(Σ\) such
that \(σ_i > σ_j\) iff \(i < j\). Doing so, the largest singular value
\(σ_0\) corresponds to the principal directions of correlation
(\(U_0\), \(V_0\)), \(σ_1\) to the second principal direction, etc.

The \(d(L,S)\) metric ranges from \(0\) 
(corresponding to perfect correlation, 
obtained for example if \(L=S\)) 
to \(n\) 
(corresponding to perfect independence), 
where $n=510$ is the number of communities considered.

To get a more precise sense of what value of \(d(L,S)\) indicates significant correlation,
We have experimentally computed \(d(L,S)\) for five random  ``embeddings'' 
with the same shape as the CCLM embeddings.
We observed a mean of \(μ_d=431.39\) and a (Bessel's-corrected) 
standard deviation \(s_d=2.90\) in their distance from the social embedding, \(S\). %\jp{Unfortunate overloading} luckily it seems that sample stdev is often written with an $s$ to differentiate from population stdev

Thus if \(d(L,S)\) is below the mean by several standard deviations, we can safely
assume that there is statistically significant correlation between
\(L\) and \(S\).  This is the case: we observe a difference between 61 and 68 standard deviations
(\cref{tab:embedding-correlations}). Further, it is clear that the CCLM embedding
capture some of the same information as the social-network embeddings.
We also give a sense of \emph{how} the correlation is manifested by
analysis of the two principal components of correlation in the
linguistic embeddings, \(U_0\) and \(U_1\). To do so we plot the
projection of each embedding along their first two principle components
which, together with the corresponding singular values, gives an idea
of how much and in what way they differ (see supplementary materials).

@@

\subsection{Results}

>>>>

We also consider the average information gain per token of the CCLM over its baseline
un-conditioned counterpart, with the same sequence encoder architecture.
For a given message, information gain is defined as the difference
between the cross-entropy of the unconditioned model and the conditioned model:
\[H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m)\]
For a set of messages, $M$, we consider the average information gain
in exponential space (as a ratio of perplexities):
\[\IG_M = \frac{e^{\avg_{m \in M}(H_{\mathrm{LM}}(m))}}{e^{\avg_{m \in M}(H_{\mathrm{CCLM}}(m))}}\]
\[\IG_M = {e^{\avg_{m \in M}(H_{\mathrm{LM}}(m) - H_{\mathrm{CCLM}}(m))}}\]
%
Unsurprisingly, the conditioned models mostly have lower perplexity 
than their respective unconditioned baseline models, 
meaning information gain above one (\cref{tab:model-results}).
Overall, the LSTM models perform better than the transformers, 
although the best transformer has higher information gain than the best LSTM.

@@

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-info-gain})\jp{missing label}.\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are \subreddit{counting}, \subreddit{friendsafari}, and \subreddit{Fireteams}, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly structured and regular formats, which 
  are presumably conventional to these communities.}
%
The variation in information gain across communities is not 
related to perplexity, however (\cref{fig:ppl-info-gain}).

@@

We observe empirically that across all the models we
tested, communities where conditioning has the least effect tend to be
more organized around more general interest topics, such as 
\subreddit{relationships} and \subreddit{advice}, where the subject
matter is relevant to a broad range of people.  Conditioning the model
on community has the most benefit for narrower special-interest
subreddits, such as those organized around a certain videogame, sports
team, or subculture.  This makes sense intuitively, since communities
with more niche topics would tend to have more specialized vocabulary
and other linguistic patterns.

@@

We can characterize the performance of the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test) set of messages $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal is clearly above the rest: the LMCC generally
picks assigns the highest probability to the community where a message originates (\cref{fig:confusion}).

@@

But we want a more quantitative measure, corresponding to the ability of the LMCC to correctly
recognize messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use a perplexity-like metric,
which is has stronger information-theoretic grounds.
To this end, we define the \emph{linguistic indiscernibility} of community $c_j$, 
noted $\Ind_j$. 
First, note that each row of $C$ is a probability distribution,
since it is computed as the average of $P(c=c_i \mid m)$ for $m\in M_j$.
The exponential of the entropy of the $j$th row of \(C\) is therefore a perplexity score,
whose maximum is the total number of communities (510 for us).
Thus we define $\Ind_j$ by scaling this perplexity by the number of communities:
\[\Ind_j = \frac{e^{-\sum_i C_{ij} log(C_{ij})}}{|C_j|}\]
This way $Ind_j=0$ means that the LMCC is completely certain of its
choice, while $Ind_j=1$ means that equal probability is given to all
predictions.  In general $\Ind_j$ is a score of uncertainty of the
LMCC, given messages from a community $c_j$.

@@

Finally, we note that, even though community conditioning affects
perplexity very little overall, it provides large benefits in term of
the underlying Bayesian classifier. Indiscernibility is on average
\num{0.57} with a standard deviation of \num{0.22}. Roughly speaking, this means
that, while the linguistic variations do not allow our model to
pinpoint the originating community of a message, it can often
eliminate TODO (a percentatge?) of communities as possible origins \bn{43\%? JP: yes}
for it. Furthermore, the information gain provided by
community-conditioning is a good predictor of how specific the
community is.\jp{By empirical observation?}

@@

The problem with this approach is that even if several dimensions of
\(L\) and \(S\) are capture similar information, they will not
coincide in the \emph{representation}. That is, applying a simple
rotation (orthogonal transformation) on either matrix widely changes
the correlation metric. To make the metric resilient to such effects,
we can compute the minimum distance between \(L_i\) and \(S_i\), for
any orthogonal matrix \(\Omega\) applied to \(L\):

@@

The problem of computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}. The
solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition (SVD) \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). That is, each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

As common when doing SVD, we can arrange \(U\), \(V\) and \(Σ\) such
that \(σ_i > σ_j\) iff \(i < j\). Doing so, the largest singular value
\(σ_0\) corresponds to the principal directions of correlation
(\(U_0\), \(V_0\)), \(σ_1\) to the second principal direction, etc.

The \(d(L,S)\) metric ranges from \(0\) (corresponding to perfect
correlation, obtained for example if \(L=S\)) to \(n\) (corresponding
to perfect orthogonality), where $n=510$ is the number of communities
considered.

To get a more precise sense of what value of \(d(L,S)\) indicates a significant correlation,
We have experimentally computed \(d(L,S)\) for ten random  ``embeddings'' 
with the same shape as the CCLM embeddings.
We observed a mean of \(μ_d=431.39\) and a (Bessel's-corrected) 
standard deviation \(s_d=2.90\) in their distance from the social embedding, \(S\). % luckily it seems that sample stdev is often written with an $s$ to differentiate from population stdev

Thus if \(d(L,S)\) is below the mean by several standard deviations, we can safely
assume that there is statistically significant correlation between
\(L\) and \(S\).  This is the case: we observe a difference between 61 and 68 standard deviations
(\cref{tab:embedding-correlations}). Further, it is clear that the CCLM embeddings
capture some of the same information as the social-network embeddings.
We also give a sense of \emph{how} the correlation is manifested by
analysis of the two principal components of correlation in the
linguistic embeddings, \(U_0\) and \(U_1\). To do so we plot the
projection of each embedding along their first two principle components
which, together with the corresponding singular values, gives an idea
of how much and in what way they differ (see supplementary materials).

@@

% \subsection{Results}
